Extracting The Power of Dependent Types

### 2021-07-08 paper.tex
Title
Commenting out fontenc
First version of section 5
Copy-paste related work and conclusion from ICFP submission (with some things removed)

\usepackage[T1]{fontenc}

@@

%% Title information
\title[]{The Power of Dependent Reflection}

@@

% Agda Basics and Extraction
\input{latex/background}
\input{latex/psembedding}
\input{latex/extraction}

@@

\section{Related Work}

\section{Conclusios and Future Work}

>>>>

%\usepackage[T1]{fontenc}

@@

%% Title information
\title[]{Extracting The Power of Dependent Types}

@@

% Agda Basics and Extraction
\input{latex/background}
\input{latex/psembedding}
\input{latex/extraction}
\input{latex/reduction}

@@

\section{\label{sec:related}Related Work}

\paragraph{Metaprogramming} There is a large body of
work on metaprogramming facilities in various programming
languages.  \citet{refl-masses} track the origins
of metaprogramming to Smith's work on reflection in Lisp~\cite{refl-lisp}.
Some prominent metaprogramming systems include
MetaOcaml~\cite{metaocaml}, MetaML~\cite{metaml},
reFlect~\cite{DBLP:journals/jfp/GrundyMO06},
Template Haskell~\cite{sheard2002template},
Racket~\cite{plt-tr1}, and various other Lisp/Scheme dialects.
%
However, these systems typically do not support dependent
types, so they are not well suited for our goal of statically
enforcing correctness of embedded programs.

\paragraph{Embedding}
Defining deep embeddings with static guarantees are a common application
of dependent types~\cite{10.5555/647849.737066,CHAPMAN200921,
10.1007/978-3-540-74464-1_7,10.1145/3236785,10.1145/1863495.1863497}.
%
These embeddings usually also define semantics of the embedded
language and therefore allow us to reason about the correctness
of program transformations and optimisations.
%
While the fact that this is possible is impressive in theory, the resulting
encodings are often difficult to use in practice. In this paper
we instead aim for a more lightweight approach.

\citet{deepshallow} propose to solve this problem with
a combination of deep and shallow embeddings.  Their idea is to define
a small deep embedding and leverage type classes in Haskell to define
the rest of the language on top of that.  It would be interesting to see
whether such an approach scales to dependently-typed embedded
languages.

\paragraph{Extraction}
The Coq proof assistant is equipped with extraction
capabilities~\cite{10.1007/978-3-540-69407-6_39,10.1007/3-540-39185-1_12},
which extracts functional code from Coq proofs (or programs).  The
default target language is Ocaml, but a few other options were added
recently.
%
Likewise, Agda itself has a mechanism for defining custom backends, of
which the GHC backend is the most prominent.
%
Other proof assistants provide similar extraction tools as well.
%
The main difference from our approach in this paper
is that these extractors are written as plugins to the proof
assistant, while we implement our extractors directly in the proof
assistant itself.
%
While it would be possible to implement extractors presented in this
paper as Coq plugins or Agda backends, conceptually they are more
heavyweight.  Our extractors and programs can (in principle)
communicate with each other. In addition, as they are just Agda programs, they can
be reflected themselves and their structure can be leveraged.

\paragraph{Dependently typed metaprogramming}
Several dependently-typed languages are equipped with metaprogramming
capabilities: Idris \cite{idris-refl}, Lean \cite{lean-refl},
Coq \cite{metacoq}, and Agda \cite{agda-refl}.  All of these
implement a similar API as described in this paper.  This is
reassuring, as it means our proposed approach is immediately portable
into many other contexts.
\citet{10.1145/3371071} introduce the Turnstile+ framework for
building dependently typed embedded DSLs and very much
shares the ideas advocated in this paper, suggesting that
our approach could work there as well.
\citet{10.1145/3371076} use MetaCoq to formally verify
the core type system of Coq. This combines nicely with
our approach, as we could use the verified core language
as a basis to verify our custom extractors.
\citet{10.1145/3372885.3373829} use MetaCoq to implement
a DSL combining deep and shallow approaches, in a way that
is quite similar to our own. While they are able to formally
reason about preservation of semantics (which we can't do yet), it is unclear
whether their approach scales to dependently-typed embedded languages.

\section{Conclusions and Future Work}

In this paper we investigate the idea of developing
embedded programs hand-in-hand with custom
code generators for them. We solve the well-known
conundrum of choosing between deep and shallow
embedding by leveraging the power of
reflection.
This allows us to enjoying the benefits
of shallow embedding, while keeping full access to the internal
structure of the embedded programs.

We apply this idea in the context of dependently-typed
languages to create verified implementations that can
be used in the context of an existing programming language.
We embed the target language in a theorem-prover, using
dependent types to encode the properties of interest.
Using reflection we bring the verified implementation
back into the original language.
%
Concretely, we have demonstrated the approach by implementing
an extractor for a fragment of the PostScript language, using
Agda as our host.

The main advantages of our approach are twofold.  First,
our solution is fine-grained --- we can chose what part
of the application to embed, and what constructs of the
host language to extract.  Secondly, our extractors
can use the full power of dependent types to guarantee
safety of our embedded programs.

Right now we cannot yet guarantee that the
extracted code preservers the semantics of the original
implementation. While we rarely see
fully-verified compiler backends in the real world,
our approach is very close to enabling this.
We would need a formal semantics of the reflected language
and the proof that reflected programs respect it.
While this is non-trivial, a system like Agda could do
this in principle.

There is a number of improvements that can be added to
Agda and our extractors to make the resulting code more
efficient.  Supporting \AK{let}s in the internal syntax
would help to preserve sharing.  Recognising irrelevance
annotations in the extractors would help to eliminate
unused function arguments.  Introducing proper language
primitives to specify what exactly is an embedding would
be helpful. And finally, having access to more of
Agda's internals such as case trees would help to
generate more performant code.

Overall, this work only scratches the surface of extraction-based
compilation.  We never considered alternative theories supported
by theorem provers, \eg{} cubical type theory in Agda; we did not
consider recursive metaprogramming; we did not consider integrating
optimisations of extracted programs other than what rewriting rules
are capable to do.  All of these offer exciting research opportunities
on the way to make verified software easily accessible.
To paraphrase Jim Morrison: ``Safety and no surprise, The End''.
